Randomized \textcolor{red}{I would prefer Stochastic} trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-24 paper/intro.tex
comments from final review pass

All methods considered in this work are based on the following simple idea: Apply an existing randomized trace estimator to $\Trace(\mtx{B}(t))$ with \emph{constant} random vectors, that is, the same randomization is used for each value of the parameter $t$. For example, the Girard-Hutchinson estimator becomes $n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$.
??? STOP HERE ???

\textcolor{red}{??? The following two paragraphs need to rewritten a bit at a later stage. It needs to be clearer what the approaches proposed in our paper are really about and the novelty wrt existing work has to come out more clearly and concretely. ??? Mention \cite{park-2024-lowrank-approximation} ?.}
\color{black}
Instead of discretizing $[a, b]$ into $t_1, \ldots, t_m$, approximating $\Trace(\mtx{B}(t_i))$ with $b(t_i)$, and analyzing the error $|\Trace(\mtx{B}(t_i)) - b(t_i)|$ for each $i$, a more natural way is to quantify the error with $\int_{a}^{b} | \Trace(\mtx{B}(t)) - b(t) |~\mathrm{d}t$, where $b(t)$ is the trace estimate in each $t \in [a, b]$. This approach has already been used in the analysis of low-rank approximations to parameter-dependent matrices \cite{kressner-2024-randomized-lowrank}.

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

\paragraph{Structure.} In \refsec{sec:analysis} we will analyze three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices. We propose and analyze an algorithm for using the aforementioned estimators to approximate the spectral density of real symmetric matrices in \refsec{sec:application}. Finally, we will test our developments on an example from electronic structure calculation in \refsec{sec:results}.

>>>>

All methods considered in this work are based on the following simple idea: Apply an existing randomized trace estimator to $\Trace(\mtx{B}(t))$ with \emph{constant} random vectors, that is, the same randomization is used for each value of the parameter $t$. For example, the Girard-Hutchinson estimator becomes $\mathsf{est}(t):=n_{\mtx{\Psi}}^{-1} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j$ for $n_{\mtx{\Psi}}$ constant Gaussian random vectors $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}}$. We will use the $L^1$-norm $\int_{a}^{b} | \Trace(\mtx{B}(t)) - \mathsf{est}(t) |~\mathrm{d}t$ to measure the error of such an estimator for all $t$ in an interval by $[a,b]$; similar to the approach taken in~\cite{kressner-2024-randomized-lowrank} for low-rank approximation.

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices with constant randomness (with respect to $t$): The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Combined with Chebyshev approximation, this allows one to reuse the majority of computations across different values of the parameter $t$, making the estimators scale favorably with respect to the number of parameter evaluations. We propose several additional modifications to the methods from \cite{lin-2017-randomized-estimation} for the special case of approximating the smoothed spectral density $\phi_{\sigma}$ from~\refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach that preserves non-negativity for approximating matrix functions in terms of Chebyshev polynomials. The resulting Chebyshev-Nyström++ estimator 
unifies all methods from~\cite{lin-2017-randomized-estimation} in a single algorithm, with the desirable side effect of \textcolor{red}{speeding up the \enquote{spectrum sweeping} method by orders of magnitude. ??? Where do we actually demonstrate this? Illustrate/spell this out more explicitly in the numerical experiments ???} Combining our analysis of the randomized trace estimators with standard error bounds for Chebyshev approximation leads to Theorem~\refthm{thm:chebyshev-nystrom}, which shows that $\mathcal O(\varepsilon^{-1})$ matrix-vector products suffice to guarantee an error $\varepsilon$.

\paragraph{Structure.} In \refsec{sec:analysis} we will analyze the error of three standard trace estimators -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with constant random matrix. In \refsec{sec:application}, we propose and analyze the Chebyshev-Nyström++ estimator, which uses the these estimators to approximate the smoothed spectral density of real symmetric matrices. Finally, we will illustrate and validate our developments for examples from various applications in \refsec{sec:results}.

### 2025-02-24 paper/numresults.tex
comments from final review pass

\color{black}

In this last section, we test the Chebyshev-Nyström++ algorithm in multiple scenarios leading from electronic structure interaction, statistical thermodynamics, and neural network optimization.

While theoretically an important tool, for the remainder of this section, we waive using the non-negative Chebyshev approximation, since the (slight) indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not seem to matter when computing the Nyström++ estimator in finite precision arithmetic. Unless otherwise mentioned, we transform the matrices such that all their eigenvalues are contained in $[-1, 1]$ by approximating the spectrum with NumPy's Hessian eigenvalue solver. We compute the spectral densities at $n_t = 100$ uniformly distributed values of the parameter $t$ in $[-1, 1]$. The integral involved in computing the $L^1$-errors is approximated with the composite midpoint quadrature rule using the $n_t$ values of $t$ as nodes. Both parameters introduced at the end \refsec{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are both set to $10^{-5}$.

All our implementations are written in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1. They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM.

@@

Our first numerical example comes from electronic structure interaction \cite[Section 6]{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
\begin{equation}
    v(r) = v_0 e^{-\lambda r^2}
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf. \reffig{fig:gaussian-well}). The grid width is fixed to be $h=0.6$. For example, for $n_c = 1$, this leads to a matrix of size $1000 \times 1000$. This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows us to interpret the system's energy levels.

\begin{figure}[ht]
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-1.pgf}
        \caption{$n_c=1$}
        \label{fig:gaussian-well-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-2.pgf}
        \caption{$n_c=2$}
        \label{fig:gaussian-well-2}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-5.pgf}
        \caption{$n_c=5$}
        \label{fig:gaussian-well-5}
    \end{subfigure}
    \caption{Cross-sections of the periodic Gaussian well potential $V$ for different sizes $n_c$ of the supercell.}
    \label{fig:gaussian-well}
\end{figure}

For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}). As a reference, we use the eigenvalues computed by NumPy's Hermitian eigenvalue solver applied to the matrix $\mtx{A}$.

\begin{figure}[ht]
    \centering
    \input{plots/convergence.pgf}
    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$, we apply the Chebyshev-Nyström++ method to the Hamiltonian matrix described in \refsec{subsec:hamiltonian}. We plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$.}
    \label{fig:convergence}
\end{figure}

Indeed, we observe that for $n_{\mtx{\Omega}} = 0$, we require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ to achieve an error of order $\varepsilon$, as shown in \refthm{thm:hutchinson}. Because the eigenvalues of the matrix are quite uniformly distributed, we observe the stronger convergence of the Nyström approximation discussed in the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is large enough.

For a fixed budget of random vectors $n_{\mtx{\Omega}} + n_{\mtx{\Psi}}$, \reffig{fig:distribution} visualizes how for different values of the smoothing parameter $\sigma$ the random vectors should either be invested into the Nyström approximation $n_{\mtx{\Omega}}$ or the estimation of the residual trace $n_{\mtx{\Psi}}$. Again, since for small $\sigma$ the matrix function $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ has a small numerical rank, a Nyström approximation alone is already enough to achieve a good approximation. On the other hand, when $\sigma$ is large, the Nyström approximation by itself is not effective, and the correction with the residual trace becomes indispensable.

\begin{figure}[ht]
    \centering
    \input{plots/distribution.pgf}
    \caption{For a range of values of $\sigma$, the Chebyshev-Nyström++ method is applied to the Hamiltonian matrix described in \refsec{subsec:hamiltonian} with $n_c = 1$ for different ways of allocating a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation. We make the approximation error committed by the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}).}
    \label{fig:distribution}
\end{figure}

@@

The Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} is also suitable in the context of spectral density estimation. It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. $\mtx{\Omega}$ is used to extract a low-rank approximation to $g_{\sigma}(t \mtx{I}_n - \mtx{A})$ by running the block-Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without. The columns of $\mtx{\Psi}$ are then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos algorithm on $\mtx{A}$.

We run our own, faithfully optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} in multiple configurations on the example from \refsec{subsec:hamiltonian} and plot the approximation errors for logarithmically spaced values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}. For reference, we add the error of the Chebyshev-Nyström++ method on parameters which lead to a comparable run-time.

\begin{figure}[ht]
    \begin{minipage}[c]{.475\linewidth}
        \centering
        \input{plots/krylov_aware_density.pgf}
    \end{minipage}\hfill%
    \begin{minipage}[c]{.475\linewidth}
        \vspace{-35pt}
        \input{tables/krylov_aware_density_KA.tex}
        \newline
        \vspace{15pt}
        \newline
        \input{tables/krylov_aware_density_CN.tex}
    \end{minipage}
    \caption{For the example from \refsec{subsec:hamiltonian} we compare the Chebyshev-Nyström++ (CN++) method with the Krylov-Aware (KA) stochastic trace estimator in multiple configurations. To this extent, we compute the $L^1$-error of the methods for various levels of smoothing $\sigma$ (left). The parameters of the algorithms and run-time averaged over all values of $\sigma$ can be found in the tables (right).}
    \label{fig:krylov-aware-density}
\end{figure}

@@

When optimizing a neural network over some data, it is of great interest to determine whether a minimum has been found and if this minimum is robust, i.e. a small perturbation of the parameters of the neural network will not lead to a significant decrease in the fit of the neural network. Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if additionally all of them are small, the minimum is robust.

Since neural networks are usually parametrized by thousands of parameters, assembling the Hessian matrix explicitly and then computing its eigenvalues is out of question. Luckily, there exists a method which can exactly compute matrix-vector products $\mtx{A} \vct{x}$ for any vector $\vct{x}$ which scales proportionally with the number of parameters \cite{pearlmutter-1994-fast-exact}.

To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters. We train this network on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.} in PyTorch 2.4.1 in a standard way. We use the bound from \cite[Theorem 1]{zhou-2011-bounding-spectrum} to estimate a bound on the spectrum.

We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss. It can well be observed that the eigenvalues creep towards zero as the training proceeds. Furthermore, despite the loss steadily decreasing, the presence of relatively large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.

\begin{figure}
    \begin{minipage}[c]{.49\linewidth}
        \centering
        \input{plots/hessian_density.pgf}
    \end{minipage}\hfill%
    \begin{minipage}[c]{.49\linewidth}
        \centering
        \input{plots/hessian_density_loss.pgf}
    \end{minipage}
    \caption{The mean squared error training loss (right) and the corresponding approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset (left). The spectral density is approximated in $n_t=150$ uniformly spaced points using the Chebyshev-Nyström++ method (\refalg{alg:nystrom-chebyshev-pp}) with parameters $n_{\mtx{\Omega}} = 10$, $n_{\mtx{\Psi}} = 10$, $m = 1000$, and $\sigma = 0.005$.}
    \label{fig:hessian-density}
\end{figure}

@@

To demonstrate that the Chebyshev-Nyström++ estimator is also effective on problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain \cite[Section 5.1]{chen-2023-krylovaware-stochastic}. We compute the pointwise error from the analytic solution as a function of the temperature $\beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.

@@

Unlike in the spectral density example from \refsec{subsec:hamiltonian}, where the numerical rank of the matrix stays low for all $t$, it changes wildly with the parameter $\beta$. The Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$. Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ is low-rank, and for small $\beta$, where the singular values only decay slowly, we cannot expect a good approximation.

>>>>

In this section, we verify the performance the Chebyshev-Nyström++ algorithm proposed in this paper in various application scenarios, from electronic structure interaction, statistical thermodynamics, and neural network optimization.

While theoretically an important tool, we have observed that the preservation of non-negativity discussed in Section~\ref{sec:preservnonneg} is not needed in practice. The (slight) indefiniteness of the standard Chebyshev approximation $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ from~\refequ{equ:matrix-approximation} does not seem to impede the reliability of the Nyström++ estimator. Unless otherwise stated, we transform the input matrices such that their eigenvalues are contained in $[-1, 1]$, by approximating their spectrum with NumPy's \textcolor{red}{Hessian ??? replace by sparse? ???} eigenvalue solver. We compute spectral densities at $n_t = 100$ uniformly distributed values of the parameter $t \in [-1, 1]$. The integral involved in computing the $L^1$-errors is approximated with the composite midpoint quadrature rule using the $n_t$ values of $t$ as nodes. Both parameters introduced at the end \refsec{subsubsec:chebyshev-nystrom-implementation},the eigenvalue truncation threshold and the parameter for detecting a vanishing spectral density, are fixed to $10^{-5}$.

Our implementations are written in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1. They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM.

@@

We use the electronic structure interaction example from~\cite[Section 6]{lin-2017-randomized-estimation}, which involves the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
%\begin{equation}
    $v(r) = v_0 e^{-\lambda r^2}$, 
%    \label{equ:5-experiments-gaussian-cell}
%\end{equation}
with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension; see \reffig{fig:gaussian-well}. The grid width of the finite difference discretization is fixed to $h=0.6$. For example, for $n_c = 1$, this leads to a matrix of size $1000 \times 1000$. This example represents an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows one to interpret the system's energy levels.

\textcolor{red}{??? This discussion of $n_c$ only makes sense if we have at least one experiment with $n_c = 5$ or so. Could you please include such an experiment? For example, do Figure 4.2 again for larger $n_c$? Just use an overapproximation (large $n_\Psi$, $n_\Omega$) for approximating the true spectral distribution. ???}

\begin{figure}[ht]
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-1.pgf}
        \caption{$n_c=1$}
        \label{fig:gaussian-well-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-2.pgf}
        \caption{$n_c=2$}
        \label{fig:gaussian-well-2}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\columnwidth}
        \input{plots/gaussian-well-5.pgf}
        \caption{$n_c=5$}
        \label{fig:gaussian-well-5}
    \end{subfigure}
    \caption{Cross-sections of the periodic Gaussian wells potential $V$ for different $n_c$.}
    \label{fig:gaussian-well}
\end{figure}

\reffig{fig:convergence} displays the convergence behavior of Chebyshev-Nyström++ applied to the $1000 \times 1000$ matrix resulting from the discretization of the Hamiltonian for $n_c = 1$. As a reference, we use the eigenvalues computed by NumPy's dense eigenvalue solver.
\begin{figure}[ht]
    \centering
    \input{plots/convergence.pgf}
    \caption{Error vs. number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ (with fixed $\sigma=0.005$ and \textcolor{red}{$m$ ??? Provide value???}) when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.}
    \label{fig:convergence}
\end{figure}%
As expected from \refthm{thm:hutchinson}, Girard-Hutchinson alone ($n_{\mtx{\Omega}} = 0$) require $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2})$ samples to achieve an error of order $\varepsilon$. Because the eigenvalues of the matrix are spread out, we observe the better convergence of the Nyström approximation discussed at the end of \refsec{sec:application} once $n_{\mtx{\Omega}}$ is sufficiently large. \textcolor{red}{??? Comment explicitly that and why the errors levels off at around $10^{-6}$. Is it because of Chebyshev or some other truncation, etc.? ???}

For a fixed budget of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}} = 80$ of random vectors, \reffig{fig:distribution} shows how the value of the smoothing parameter $\sigma$ the choice between Nyström ($n_{\mtx{\Omega}}$) and 
Girard-Hutchinson ($n_{\mtx{\Psi}}$). As expected, for small $\sigma$ the Nyström approximation alone suffices, but 
this is not an effective choice for large $\sigma$. Choosing $n_{\mtx{\Psi}} = n_{\mtx{\Omega}}$ constitutes a good compromise.

\begin{figure}[ht]
    \centering
    \input{plots/distribution.pgf}
    \caption{Error vs. $\sigma$ for a fixed budget of $80$ random vectors 
    when applying  Chebyshev-Nyström++ to the Hamiltonian matrix from~\refsec{subsec:hamiltonian} with $n_c = 1$.
The Chebyshev approximation error is made negligible by setting $m=16 / \sigma$ (see \reflem{lem:non-negative-chebyshev-error}).}
    \label{fig:distribution}
\end{figure}

@@

The Krylov-aware stochastic trace estimator from \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} is a viable alternative method for spectral density estimation. \textcolor{red}{??? Please rewrite this. It does not come out whether~\cite{chen-2023-krylovaware-stochastic} actually cover spectral density estimation. Are the Gaussian matrices chosen constants in their paper/in our experiments? If this is not discussed in~\cite{chen-2023-krylovaware-stochastic}, comment more explicitly to which extent this algorithm benefits from constant random matrices. Do they have an analysis that applies to our situation, etc. ??? It also samples two Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. $\mtx{\Omega}$ is used to extract a low-rank approximation to $g_{\sigma}(t \mtx{I}_n - \mtx{A})$ by running the block-Lanczos algorithm on $\mtx{A}$ with starting block $\mtx{\Omega}$ for $r$ iterations with reorthogonalization and subsequently $q$ without. The columns of $\mtx{\Psi}$ are then used to estimate the trace of the approximation residual with $r$ iterations of the Lanczos algorithm on $\mtx{A}$ ??? Why $\mtx{A}$? ???.}

We run our own, faithfully optimized implementation of the Krylov-aware stochastic trace estimator \cite[Algorithm 3.1]{chen-2023-krylovaware-stochastic} for different parameter settings on the example from \refsec{subsec:hamiltonian} and plot the approximation errors for logarithmically spaced values of the smoothing parameter $\sigma$ in \reffig{fig:krylov-aware-density}. For reference, we also report the error of Chebyshev-Nyström++ for parameters that lead to a comparable run-time. While the the Krylov-aware estimator is clearly an attractive alternative, it requires a careful choice of parameters adapted to  different regimes of $\sigma$ and still leads to a significantly larger error for moderately small values $\sigma$.

\begin{figure}[ht]
    \begin{minipage}[c]{.475\linewidth}
        \centering
        \input{plots/krylov_aware_density.pgf}
    \end{minipage}\hfill%
    \begin{minipage}[c]{.475\linewidth}
        \vspace{-35pt}
        \input{tables/krylov_aware_density_KA.tex}
        \newline
        \vspace{15pt}
        \newline
        \input{tables/krylov_aware_density_CN.tex}
    \end{minipage}
    \caption{For the example from \refsec{subsec:hamiltonian}, error vs. $\sigma$ the Krylov-Aware (KA) stochastic trace estimator with different parameter settings, compared to the Chebyshev-Nyström++ (CN++) method.}
    \label{fig:krylov-aware-density}
\end{figure}

@@

When fitting the parameters of a neural network, it can be of interest to determine whether a (local) minimum has been found and whether this minimum is robust, i.e., whether a small perturbation of the parameters leads to a significant decrease in the fit of the neural network. Both properties are reflected in the eigenvalues of the Hessian matrix $\mtx{A}$ with respect to some loss function: if all eigenvalues are non-negative, then the loss attains a local minimum, and if, additionally, all of them are small, the minimum is robust. \textcolor{red}{??? Isn't it the opposite? If the eigenvalues are not close to zero, we get a robust local minimum? Please check! ???}

Since neural networks are usually parametrized by a large number of parameters, assembling the Hessian matrix explicitly and then computing its eigenvalues is clearly infeasible. On the other hand, one can cheaply compute matrix-vector products $\mtx{A} \vct{x}$ with the Hessian, at a cost that scales proportionally with the number of parameters \cite{pearlmutter-1994-fast-exact}.

To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with $6\,782$ parameters. We train this network on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.} in PyTorch 2.4.1 in a standard way. We use the bound from \cite[Theorem 1]{zhou-2011-bounding-spectrum} to estimate bounds on the spectrum.

We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss. It can be observed that the eigenvalues creep towards zero as the training proceeds. Furthermore, despite the loss steadily decreasing, the presence of relatively large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.

\begin{figure}
    \begin{minipage}[c]{.49\linewidth}
        \centering
        \input{plots/hessian_density.pgf}
    \end{minipage}\hfill%
    \begin{minipage}[c]{.49\linewidth}
        \centering
        \input{plots/hessian_density_loss.pgf}
    \end{minipage}
    \caption{Mean squared error training loss (right) and corresponding approximate spectral density of the Hessian matrix of a fully connected convolutional neural network in different epochs of training on the MNIST dataset (left). The spectral density is approximated in $n_t=150$ uniformly spaced points using Chebyshev-Nyström++ with parameters $n_{\mtx{\Omega}} = 10$, $n_{\mtx{\Psi}} = 10$, $m = 1000$, and $\sigma = 0.005$.}
    \label{fig:hessian-density}
\end{figure}

@@

To demonstrate that the Chebyshev-Nyström++ estimator is also effective for problems unrelated to spectral density estimation, we consider the problem of approximating the partition function $\Trace(\exp(-\beta \mtx{A}))$ for the Hamiltonian $\mtx{A}$ of a planar Heisenberg spin chain from~\cite[Section 5.1]{chen-2023-krylovaware-stochastic}. We compute the pointwise error from the analytic solution as a function of the temperature $t\equiv \beta^{-1}$ and compare it to the error achieved by the Krylov-aware trace estimator in \reffig{fig:krylov-aware-spin}.

@@

Unlike for the spectral density example from \refsec{subsec:hamiltonian}, where the numerical rank of the matrix is relatively uniform across different $t$, this now changes changes wildly with the parameter $\beta$. Recall that the Chebyshev-Nyström++ method requires a fixed allocation of random vectors to the low-rank approximation for all values of the parameter $\beta$. Hence, both for large $\beta$, where the matrix $\exp(-\beta \mtx{A})$ has low numerical rank, and for small $\beta$, where the singular values decay very slowly, the approximation error is relatively large.
\textcolor{red}{??? Could you please recheck this example? Why is CN++ extremely bad for small $\beta^{-1}$? and KA much better? Please comment on this aspect explicitly. ???}

### 2025-02-24 paper/paper.tex
review comments
adapt nomenclature and fix README
remove green color
review Chebyshev coefficient computation
polish results and add conclusion
comments from final review pass

\documentclass[12pt]{article}

@@

\title{Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation}

@@

\begin{abstract}
    Stochastic trace estimators are a family of well established techniques for approximating the trace of constant matrices. Oftentimes the matrix whose trace we aim to approximate continuously depends on a real parameter $t \in [a,b]$. This is for example the case when approximating the spectral density of a matrix or computing partition functions of quantum systems. Clearly, we could separately approximate the traces of the constant matrices $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ at the fixed values $t_1, \dots, t_m \in [a, b]$. However, this approach does not scale well if $m$ is large. We analyze three trace estimators for constant matrices -- the Girard-Hutchinson, Nyström, and Nyström++ estimators -- when they are applied to parameter-dependent matrices with the same randomization reused for all values of the parameter $t$. We show that all of these estimators closely match their corresponding error guarantees for constant matrices. In particular, for the Nyström++ method, we can guarantee an error of at most $\varepsilon$ by computing the estimate with just $\mathcal{O}(\varepsilon^{-1})$ random vectors, independent of what structure the symmetric positive semi-definite matrix $\mtx{B}(t)$ has. We employ these trace estimators to improve upon a technique for approximating the spectral density of a matrix and provide a detailed analysis of the resulting method. Finally, we confirm our theoretical developments in multiple numerical experiments.
\end{abstract}

\textcolor{red}{TODO for Fabio:
\begin{itemize}
%\item Add author list. Matti first, rest (He/Kressner/Lam) in alphabetical order.
 \item The text style is a bit too generous but no need to change this now. \textcolor{green}{Ok! Let's talk about it at some point.}
\end{itemize}
}
\color{blue}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
\input{numresults.tex}

%\section{Conclusion}
%\label{sec:conclusion}
%
%\todo{
%\begin{itemize}
%    \item Robustness (can choose $m$ and $n_{\mtx{\Omega}}$, $n_{\mtx{\Psi}}$ as large as one wants, and not break algorithm, unlike Krylov-aware Lanczos)
%\end{itemize}
%}
\paragraph{Acknowledgements.} The authors gratefully acknowledge the help of Lin Lin in reproducing the example in \refsec{subsec:hamiltonian}. They thank David Persson for many enlightening discussions relating the proof in \refsec{subsec:nystrom-pp}.

>>>>

\documentclass[11pt]{article}

@@

\title{Randomized \textcolor{red}{I would prefer Stochastic} trace estimation for parameter-dependent matrices applied to spectral density approximation}

@@

\begin{abstract}
    Stochastic trace estimation is a well-established tool for approximating the trace of a large symmetric positive semi-definite matrix $\mtx{B}$. Several applications involve a matrix that depends continuously on parameter $t \in [a,b]$, and require trace estimates of $\mtx{B}(t)$ for many different values of $t$. This is, for example, the case when approximating the spectral density of a matrix or computing partition functions of quantum systems. Approximating the trace separately for each matrix
    $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ clearly incurs redundancies and a cost that scales linearly with $m$. To address this issue, we propose and analyze modifications for three stochastic trace estimators, the Girard-Hutchinson, Nyström, and Nyström++ estimators. Our modification uses \emph{constant} randomization across different values of $t$, that is,
    every matrix $\mtx{B}(t_1), \dots, \mtx{B}(t_m)$ is multiplied with the \emph{same} set of random vectors.
    When combined with Chebyshev approximation in $t$, the use of such constant random matrices allows one to reuse matrix vector-products across different values of $t$, leading to significant cost reduction.
    Our analysis shows that the loss of stochastic independence across different $t$ does not lead to detoriation; in fact, our error bounds for these estimators closely match existing error bounds for the constant case. In particular, we show that $\mathcal{O}(\varepsilon^{-1})$ random matrix-vector products suffice to ensure an error of $\varepsilon > 0$ for Nyström++, independent of low-rank properties of $\mtx{B}(t)$. We discuss in detail how the combination of Nyström++ with 
    Chebyshev approximation applies to large-scale spectral density estimation and provide an analysis of the resulting method. This  improves various aspects of an existing stochastic estimator for spectral density estimation. Several numerical experiments for examples from electronic structure interaction, statistical thermodynamics, and neural network optimization illustrate and validate our findings.
\end{abstract}

\textcolor{red}{TODO for Fabio:
\begin{itemize}
 \item Has~\cite{bhattacharjee-2024-improved-spectral} been published?
 \item Fix discussion end of Section 3.3.2
 \item Please reread entire Section 4 and address various red points.
 \item Please reread abstract.
 \item Please reread last part of Section 2 (startin from new contributions) and fix red remark.
 \item Please remove all unreferenced equation labels (using checkref) at the end.
 \item Perform a spell check at the end.
\end{itemize}
}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
\input{numresults.tex}
\input{conclusion.tex}

\paragraph{Acknowledgements.} The authors gratefully acknowledge the help of Lin Lin in reproducing the example in \refsec{subsec:hamiltonian}. They thank David Persson for many enlightening discussions relating the proof in \refsec{subsec:nystrom-pp}.
